Formal Modeling and Verification of a Rate-Monotonic Scheduling Implementation with Real-Time Maude

### 2015-11-06 RMS/submission.tex
almost final version, to be sent to Prof.Song.

\usepackage{alltt}
%\usepackage{epic}
%\usepackage{pstricks}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\newcommand{\hide}[1]{\ignorespaces}
\newcommand{\jx}[1]{{\bf Jiaxiang: }#1{ \bf End}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}

@@

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
\begin{abstract}
The abstract goes here. TODO.
\end{abstract}

@@

It is even more crucial to ensure the reliability of an implementation
instead of the algorithm, when RMS serves in a safety-critical
system. When analyzing a realistic implementation, theoretical results
may be no more applicable. For instance, even though the conditions
derived from algorithm analysis are satisfied, schedulability can be
broken by overhead in the system, or by the interrupt masking
mechanism which may delay interrupt handling. On the other hand,
correctness of the implementation with respect to the algorithm is
difficult to be verified by the traditional methods such as testing
and simulation, since they are incomplete. Extensive effort to apply
formal methods, such as model checking and theorem proving, has been
made to analyze safety-critical systems for the past few
years~\cite{DBLP:journals/tie/MiyawakiMSYV05,DBLP:journals/iandc/MeseguerR13,DBLP:journals/cacm/Leroy09,DBLP:conf/sosp/KleinEHACDEEKNSTW09}. However,
as far as we know, few~\cite{DBLP:conf/iceccs/CuiDT14,TianD2011}
attempt to analyze the RMS algorithm, while no work for
implementations of RMS is found.

@@

A simple example showing this algorithm is depicted in
Figure~\ref{sf:rmsalg}. However in this paper, we consider an
implementation instead of the RMS algorithm itself, thus the model
would be more complicated than this standard, ideal
setting. Assumption~(A1) will be modified because of the interrupt
masking mechanism, while (A2) and (A4) are relaxed to obtain a more
realistic analysis model.

@@

\section{The Implementation of RMS}
\label{s:imp}
The implementation is in a realistic avionic control system.
Interrupts would be triggered by, and only by, the clock every $T$,
which we call \emph{interrupt cycle}. When an interrupt request
occurs, if the system is interruptable, i.e. the interrupt masking bit
is cleared, the handler function $schedule()$ will be invoked;
otherwise $schedule()$ will be pending until the interrupt masking bit
becomes cleared.  The implementation is shown as $schedule()$ in
Figure~\ref{a:schedule}, where $taskList$ is the list of periodic
tasks to be scheduled. We assume that the list is in descending order
of priority, and both variables $taskList$ and $timer$ are global. In
this implementation, there is only one kind of interrupt, the period
$T_i$ of each task is a multiple of $T$, and the tasks are
independent, meeting assumption~(A3).

@@

In Figure~\ref{a:schedule}, the handler function $scheduler()$ first
updates status of all tasks in $taskList$ via function
$updateStatus()$. This updating actually initiates tasks which should
be scheduled in the current interrupt cycle. Then $scheduler()$
traverses the list to execute the ready tasks one by one, or to do a
return when encountering an interrupted\footnote{Note that the status
  \textit{INTERRUPT} indicates the task is interrupted for the moment,
  or was interrupted before but its execution is not complete yet.}
task. As for the function $updateStatus()$, it updates each task in
two steps: firstly, if the task is running, it becomes interrupted;
secondly for the task at its initiation time, if its previous job is
complete, it would be set to be ready, otherwise it misses its
deadline, producing an error. Notice that $schedule()$ is invoked only
when the interrupt request is handled, not when the interrupt is
disabled (the interrupt masking bit is set). Due to the interrupt
masking bit, its execution cannot be interrupted when it is updating
status of tasks or finding the next task to execute, however, it can
be interrupted while executing some task (Line~\ref{l:function}). This
allows the execution of $schedule()$ to be nested.

@@

\section{Formal Modeling of the Implementation}
\label{s:formalism}
Considering some technical details of the platform, such as interrupt
masking mechanism, we model the implementation under the following
assumptions:

@@

These assumptions make our model different from the standard one. For
instance, with assumption~(A1'), an interrupt request that occurs
while switching is performed will be pending, such that jobs of task
$\tau_i$ cannot initiate at $kT_i$. They should wait until switching
finishes and the interrupt masking bit is cleared, which is different
from~(A1). Note that (A3) says that jobs are ready to run at their
initiation times, however, no one can start running at exactly its
initiation time, because scheduling takes time. Under these
assumptions, an example for showing the execution of tasks scheduled
by the implementation is depicted in Figure~\ref{sf:rmsimp}. Note that
the second job of $\tau_1$ is initiated at time $5.5$ instead of $5$,
because switching is performed and the interrupt masking bit is set at
time $5$.

@@

In this section, we introduce first how we model the states--the
static aspect--of the system using algebraic data types, then how we
specify the essential behaviors--the dynamic aspect--using rewrite
rules. In particular, modeling of instantaneous behaviors would be
explained in Sections~\ref{ss:ir}, \ref{ss:init}
and~\ref{ss:inthandling}, followed by timed behaviors in
Section~\ref{ss:timedbehavior}.

@@

A sort \verb|Stack| is introduced to simulate the stack of the system,
storing the tasks that are being interrupted, and equipped with
operations \verb|push|, \verb|pop| and \verb|peek| on it.
\begin{verbatim}
  op bottom : -> Stack [ctor] .
  op _#_ : Nat Stack -> Stack [ctor] .
\end{verbatim}

@@

A periodic task, which is needed to be scheduled, is an object
instance of the subclass \verb|PTask| with additional attributes
\verb|priority|, \verb|period| and \verb|status|:
\begin{verbatim}
  class PTask | priority : Nat, 
                period : NzNat, 
                status : Status .
  subclass PTask < Task .
\end{verbatim}
where \verb|Status| is a sort with four constant constructors, same as
in the implementation:
\begin{verbatim}
  ops RUNNING INTERRUPT READY DORMANT 
        : -> Status [ctor] .
\end{verbatim}
The list of periodic tasks, the variable $taskList$ in the
implementation, is modeled as an instance of the sort \verb|TaskList|,
which is a list of \verb|PTask|s\footnote{Following the Maude
  convention, variables would be written in capital letters.}:
\begin{verbatim}
  op null : -> TaskList [ctor] .
  op _::_ : Object TaskList 
              ~> TaskList [ctor] .
  mb (< O:Oid : PTask |> :: L:TaskList) 
       : TaskList .
  mb (error :: L:TaskList) : TaskList .
\end{verbatim}
Periodic tasks are identified by their indexes in the list.

@@

The set of registers is modeled as an object instance of the class
\verb|Regs|, with attributes \verb|pc| denoting the program counter,
\verb|mask| for the interrupt masking flag and \verb|ir| for the
interrupt request flag, respectively.
\begin{verbatim}
  class Regs | pc : TaskID, 
               mask : Bool, ir : Bool .
\end{verbatim}
where the sort \verb|TaskID| is a supersort of \verb|MaybeNat| and
\verb|Oid|, referring to some task.
\begin{verbatim}
  subsorts MaybeNat Oid < TaskID .
\end{verbatim}
Some operations, such as \verb|getPc| and \verb|setMask|, are defined
on the class \verb|Regs|.

Then the hardware is naturally described by the sort \verb|Hardware|:
\begin{verbatim}
  op [_;_] : Object Stack 
               ~> Hardware [ctor] .
  mb ([ < O:Oid : Regs |> ; S:Stack ]) 
       : Hardware .
\end{verbatim}

@@

\subsubsection{System}
The entire system in our model is a composition of the parts
introduced above, with a sort \verb|System|\footnote{Some variable
  declarations are not shown for simplicity.}:
\begin{verbatim}
  op _____ : TaskList Timer 
             SysTasks Hardware Object 
               ~> System [ctor] .
  mb (L T STS HW < O:Oid : IntSrc |>) 
       : System .
\end{verbatim}

\subsection{Interrupt Requests}
\label{ss:ir}
Interrupt requests are performed by the source exactly every cycle
$T$, when the attribute \verb|val| decreases to \verb|zero|. The
requesting is an instantaneous action, thus is modeled by the
following instantaneous conditional rule applied on \verb|System|:
\begin{verbatim}
  crl [interrupt-request] :
    (L T STS HW ISRC) 
    => (L T STS (HW).intReq reset(ISRC))
    if (ISRC).timeout .
\end{verbatim}
where the function \verb|_.timeout| examines whether the attribute
\verb|val| equals \verb|zero|, and \verb|_.intReq| sets the \verb|ir|
flag indicating there exists an interrupt request to be handled:
\begin{verbatim}
  op _.intReq : Hardware -> Hardware .
  eq [ REGS ; S ].intReq 
     = [ (REGS).setIr ; S ] .
\end{verbatim}
Then the request will wait to be handled, which is explained in
Section~\ref{ss:inthandling}.

@@

\subsection{Interrupt Handling and Task Scheduling}
\label{ss:inthandling}
When an interrupt request occurs, it may not be detected immediately
by the system. It requires the \verb|mask| flag to be cleared. Once
the request is detected, it is handled by two steps: the interrupt
handling mechanism of the hardware (such as clearing \verb|ir|,
pushing context into stack and so on), and to invoke the $schedule()$
function. This behavior is modeled by the following instantaneous
rewrite rule:
\begin{verbatim}
  crl [interrupt-handle] :
    SYSTEM 
    => ((SYSTEM).interrupt).startScheduling
    if (SYSTEM).existInt .
\end{verbatim}
where \verb|_.existInt| checks whether \verb|mask| is cleared
\emph{and} \verb|ir| is set. The function \verb|_.interrupt| models
the interrupt handling mechanism performed by the hardware and does
four things: (i) clearing the \verb|ir| flag, which means the request
has been handled; (ii) pushing the current \verb|pc| value into the
stack, storing the interrupted context; (iii) assigning
\verb|scheduling| of sort \verb|Oid| to \verb|pc|, which indicates
that the system is scheduling; and (iv) setting the \verb|mask| flag,
to mask coming interrupt requests.

Unlike periodic tasks, even though the scheduling stage is modeled by
a \verb|Counter|, its functionality is too important to be abandoned.
We divide the behaviors of scheduling into three parts.  The first
part contains its timed behaviors. This part is modeled by regarding
scheduling as a system task of sort \verb|SysTask|. Modeling timed
behaviors of tasks is explained in Section~\ref{ss:timedbehavior}.
The other two parts together define its functionality. The second part
corresponds to Lines~\ref{l:updatestatus}-\ref{l:inc} in
Figure~\ref{a:schedule}. It performs updating the status of
$taskList$ and increasing $timer$ by $1$. This part is modeled by
function \verb|_.startScheduling| which applies instantaneously at the
beginning of scheduling, as shown in rule \verb|interrupt-handle|:
\begin{verbatim}
  op _.startScheduling : System -> System .
  eq (L T STS HW ISRC).startScheduling 
     = ((updateStatus L with T) 
        inc(T) STS HW ISRC) .
\end{verbatim}
The third part corresponds to
Lines~\ref{l:startrun1st}-\ref{l:endrun1st}, finding the first
should-be-run periodic task and setting it to execute. It is modeled
by function \verb|_.finishScheduling|, which applies instantaneously
at the end of scheduling:
\begin{verbatim}
  op _.finishScheduling : System -> System .
  eq (L T STS HW ISRC).finishScheduling
     = (L T (finish scheduling in STS) 
        HW ISRC).run1stTask .
\end{verbatim}
where \verb|finish_in_| resets the counter of task \verb|scheduling|,
and \verb|_.run1stTask| models
Lines~\ref{l:startrun1st}-\ref{l:endrun1st} in
Figure~\ref{a:schedule}, searching the task with highest priority
that has status \verb|INTERRUPT| or \verb|READY| then performing an
\emph{interrupt return} or executing it, respectively.

@@

Similar to scheduling, the switching stage is divided into timed
behaviors of \verb|switching| and its functionality. \verb|switching|
starts when the execution time of the running periodic task reaches
its computation requirement, and finishes when its own execution time
does so. Two similar instantaneous rules are defined:
\begin{verbatim}
  crl [task-finish] :
    (L T STS HW ISRC) 
    => (SYSTEM).startSwitching
    if SYSTEM := (L T STS HW ISRC)
       /\ some N := (SYSTEM).running
       /\ some N mayFinish?in L .
  crl [switching-finish] :
    (L T STS HW ISRC) 
    => (SYSTEM).finishSwitching
    if SYSTEM := (L T STS HW ISRC)
       /\ (SYSTEM).running == switching
       /\ switching mayFinish?in STS .
\end{verbatim}

\subsection{Timed Behaviors of the System}
\label{ss:timedbehavior}
\emph{Timed behaviors} of the system consist of two parts: the
execution of tasks and the execution of the interrupt source. Both are
modeled together by the following single ``standard'' tick
rule~\cite{DBLP:journals/entcs/OlveczkyM07a}:
\begin{verbatim}
  crl [tick]:
    {SYSTEM} => {delta(SYSTEM, R)} in time R 
    if R le mte(SYSTEM) [nonexec] .
\end{verbatim}
where \verb|delta| defines the effects of time elapse on the system,
and \verb|mte| denotes the \emph{m}aximum amount of \emph{t}ime
allowed to \emph{e}lapse from the current state until an instantaneous
transition \emph{must} be performed. In fact, the core to model timed
behaviors is to define functions \verb|delta| and \verb|mte|. Notice
that the variable $R$ is \emph{continuous} with respect to the
specific time domain\footnote{Real-Time Maude contains built-in
  modules to define the time domain to be natural numbers and rational
  numbers, specifying \emph{discrete} time domains and \emph{dense}
  time domains, respectively.}  that we choose to instantiate our
model on, which is different from timed automata that discretize dense
time by defining ``clock region''.

@@

\verb|mte|, the maximum amount of time allowed to elapse, depends on
when the next instantaneous action must perform. Therefore, it is
decided by three arguments: the remaining time to finish the running
task, the remaining time to request the next interrupt, and whether or
not there exists an interrupt request detected for the moment:
\begin{verbatim}
  ceq mte(L T STS HW ISRC)
      = minimum( mteTask(ID, L),
                 minimum( mteIS(ISRC), 
                          mteIr(HW)))
      if ID := (HW).getPc 
         /\ ID :: MaybeNat .
\end{verbatim}
where \verb|mteIr| returns \verb|zero| if there exists an interrupt
request detected in the system, or \verb|INF| which represents
\emph{infinity} otherwise. Again we do not show the case where
\verb|ID| is of sort \verb|Oid|, which is very similar. We should
point out that \verb|mteTask| computes the remaining time to reach the
maximum of the computation requirement of the task, since that is the
time at which a \verb|task-finish| transition \emph{must} happen.

@@

\subsection{Properties}
The main objective is to verify the schedulability of a given set of
periodic tasks. To this end, we define an atomic proposition
\verb|taskTimeout| to hold if there exists an \verb|error| appearing
in the $taskList$ of the current state:
\begin{verbatim}
  op taskTimeout : -> Prop [ctor] .
  eq {L T STS HW ISRC} |= taskTimeout 
     = containError(L) .
\end{verbatim}
where \verb|containError| returns \verb|true| if there is an object
\verb|error| existing in $L$. Then our desired
property--schedulability--can be formalized as the temporal logic
formula: \verb|[](~taskTimeout)|. As the property is not
\emph{clocked}, given an initial state \verb|init|, the following
untimed model checking command returns \verb|true| if the
schedulability property holds with no time limit; otherwise a trace
showing a counterexample is provided:
\begin{verbatim}
  (mc init |=u [](~taskTimeout) .)
\end{verbatim}

Another objective is to verify the correctness of the implementation,
in the sense that the periodic tasks are scheduled exactly with
respect to their priorities. The atomic proposition
\verb|scheduleCorrectly| is hence defined to hold if the running
periodic task is the one requested to be executed with the highest
priority:
\begin{verbatim}
  op scheduleCorrectly : -> Prop [ctor] .
  ceq {L T STS HW ISRC} |= scheduleCorrectly 
      = (not containError(L))
        and (if ID :: MaybeNat 
             then shouldRun(ID, L)
             else true fi)
      if ID := (HW).getPc .
\end{verbatim}
where \verb|shouldRun(ID, L)| returns \verb|true| if the task
identified by \verb|ID|, probably \verb|none|, is the one possessing
the highest priority among those whose status is \verb|READY|,
\verb|RUNNING| or \verb|INTERRUPT|. Note that \verb|scheduleCorrectly|
requires no \verb|error| popping up, since our defined correctness
property makes no sense in the implementation once some task misses
its deadline. The correctness property is formalized as the temporal
logic formula: \verb|[]scheduleCorrectly|, and can be verified by the
following untimed model checking command provided an initial state
\verb|init|:
\begin{verbatim}
  (mc init |=u []scheduleCorrectly .)
\end{verbatim}

\subsection{Scenarios}
\label{ss:results}
We use the following setting for our verification, which is from the 
statistics provided by our industrial partner:
\begin{itemize}
\item The interrupt cycle is $5ms$.
\item The scheduling time ranges from $15{\mu}s$ to $38{\mu}s$, and the
switching time ranges from $10{\mu}s$ to $20{\mu}s$.
\hide{\item The scheduling time ranges from $5{\mu}s$ to $9{\mu}s$, and the
switching time ranges from $2{\mu}s$ to $4{\mu}s$.}
\item The initial state is with empty stack, empty \verb|pc|, cleared 
\verb|mask| and cleared \verb|ir|.
\end{itemize}

>>>>

\usepackage{alltt}
%\usepackage{epic}
%\usepackage{pstricks}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\newcommand{\hide}[1]{\ignorespaces}
\newcommand{\jx}[1]{{\bf Jiaxiang: }#1{ \bf End}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

@@

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
\begin{abstract}
Rate-Monotonic Scheduling (RMS) is one of the most important real-time
scheduling algorithms. There are large numbers of results about RMS,
especially on its schedulability. However, these theoretical results
contain not enough details to be used directly to investigate a
realistic RMS implementation. On the other hand, the correctness of
such an implementation is of the same importance. In this paper, we
analyze a realistic RMS implementation by formal modeling using
Real-Time Maude, a language and tool based on rewriting
logic. Overhead and some details of the hardware are taken into
account in the model. We verify the schedulability and the correctness
of the implementation within several scenarios, using model checking
tool built in Real-Time Maude. The soundness and the completeness of
our results are proved in the end.
\end{abstract}

@@

It is even more crucial to ensure the reliability of an implementation
instead of the algorithm, when RMS serves in a safety-critical
system. When analyzing a realistic implementation, theoretical results
may be no more applicable. For instance, even though the conditions
derived from algorithm analysis are satisfied, schedulability can be
broken by overhead in the system, or by the interrupt mask mechanism
which may delay interrupt handling. On the other hand, correctness of
the implementation with respect to the algorithm is difficult to be
verified by the traditional methods such as testing and simulation due
to their incompleteness. Extensive effort to apply formal methods,
such as model checking and theorem proving, has been made to analyze
safety-critical systems for the past few
years~\cite{DBLP:journals/tie/MiyawakiMSYV05,DBLP:journals/iandc/MeseguerR13,DBLP:journals/cacm/Leroy09,DBLP:conf/sosp/KleinEHACDEEKNSTW09}. However,
as far as we know, few~\cite{DBLP:conf/iceccs/CuiDT14,TianD2011}
attempt to analyze the RMS algorithm, while no work for
implementations of RMS is found.

@@

A simple example showing this algorithm is depicted in
Figure~\ref{sf:rmsalg}. However in this paper, we consider an
implementation instead of the RMS algorithm itself, thus the model
would be more complicated than this standard, ideal
setting. Assumption~(A1) will be modified because of the interrupt
mask mechanism, while (A2) and (A4) are relaxed to obtain a more
realistic analysis model.

@@

\section{The Implementation of RMS}
\label{s:imp}
The implementation is in a realistic avionic control system.
Interrupts would be triggered by, and only by, the clock every $T$,
which we call \emph{interrupt cycle}. When an interrupt request
occurs, if the system is interruptable, i.e. the interrupt mask bit
is cleared, the handler function $schedule()$ will be invoked;
otherwise $schedule()$ will be pending until the interrupt mask bit
becomes cleared.  The implementation is shown as $schedule()$ in
Figure~\ref{a:schedule}, where $taskList$ is the list of periodic
tasks to be scheduled. We assume that the list is in descending order
of priority, and both variables $taskList$ and $timer$ are global. In
this implementation, there is only one kind of interrupt, the period
$T_i$ of each task is a multiple of $T$, and the tasks are
independent, meeting assumption~(A3).

@@

In Figure~\ref{a:schedule}, the handler function $schedule()$ first
updates status of all tasks in $taskList$ via function
$updateStatus()$. This updating actually initiates tasks which should
be scheduled in the current interrupt cycle. Then $schedule()$
traverses the list to execute the ready tasks one by one, or to do a
return when encountering an interrupted\footnote{Note that the status
  \textit{INTERRUPT} indicates the task is interrupted for the moment,
  or was interrupted before but its execution is not complete yet.}
task. As for the function $updateStatus()$, it updates each task in
two steps: firstly, if the task is running, it becomes interrupted;
secondly for the task at its initiation time, if its previous job is
complete, it would be set to be ready, otherwise it misses its
deadline, producing an error. Notice that $schedule()$ is invoked only
when the interrupt request is handled, not when the interrupt is
disabled (the interrupt mask bit is set). Due to the interrupt
mask bit, its execution cannot be interrupted when it is updating
status of tasks or finding the next task to execute, however, it can
be interrupted while executing some task (Line~\ref{l:function}). This
allows the execution of $schedule()$ to be nested.

@@

\section{Formal Modeling of the Implementation}
\label{s:formalism}
Considering some technical details of the platform, such as interrupt
mask mechanism, we model the implementation under the following
assumptions:

@@

These assumptions make our model different from the standard one. For
instance, with assumption~(A1'), an interrupt request that occurs
while switching is performed will be pending, such that jobs of task
$\tau_i$ cannot initiate at $kT_i$. They should wait until switching
finishes and the interrupt mask bit is cleared, which is different
from~(A1). Note that (A3) says that jobs are ready to run at their
initiation times, however, no one can start running at exactly its
initiation time, because scheduling takes time. Under these
assumptions, an example for showing the execution of tasks scheduled
by the implementation is depicted in Figure~\ref{sf:rmsimp}. Note that
the second job of $\tau_1$ is initiated at time $5.5$ instead of $5$,
because switching is performed and the interrupt mask bit is set at
time $5$.

@@

In this section, we introduce first how we model the states--the
static aspect--of the system using algebraic data types, then how we
specify the essential behaviors--the dynamic aspect--using rewrite
rules. In particular, modeling of \emph{instantaneous behaviors} would
be explained in Sections~\ref{ss:ir}, \ref{ss:init}
and~\ref{ss:inthandling}, followed by \emph{timed behaviors} in
Section~\ref{ss:timedbehavior}.

@@

A sort \verb|Stack| is introduced to model the stack of the system,
storing the tasks that are being interrupted, and equipped with
operations \verb|push|, \verb|pop| and \verb|peek| on it.
\begin{verbatim}
 op bottom : -> Stack [ctor] .
 op _#_ : Nat Stack -> Stack [ctor] .
\end{verbatim}

@@

A periodic task, which needs to be scheduled, is an object instance of
the subclass \verb|PTask| of \verb|Task| with additional attributes
\verb|priority|, \verb|period| and \verb|status|:
\begin{verbatim}
 class PTask | priority : Nat, 
               period : NzNat, 
               status : Status .
 subclass PTask < Task .
\end{verbatim}
where \verb|Status| is a sort with four constant constructors, same as
in the implementation:
\begin{verbatim}
 ops RUNNING INTERRUPT READY DORMANT 
       : -> Status [ctor] .
\end{verbatim}
The list of periodic tasks, the variable $taskList$ in the
implementation, is modeled as an instance of the sort \verb|TaskList|,
which is a list of \verb|PTask|s\footnote{Following the Maude
  convention, variables would be written in capital letters.}:
\begin{verbatim}
 op null : -> TaskList [ctor] .
 op _::_ : Object TaskList 
             ~> TaskList [ctor] .
 mb (< O:Oid : PTask |> :: L:TaskList) 
      : TaskList .
 mb (error :: L:TaskList) : TaskList .
\end{verbatim}
Periodic tasks are identified by their indexes in the list.

@@

The set of registers is modeled as an object instance of the class
\verb|Regs|, with attributes \verb|pc| denoting the program counter,
\verb|mask| for the interrupt mask bit and \verb|ir| for the interrupt
request bit, respectively.
\begin{verbatim}
 class Regs | pc : TaskID, 
              mask : Bool, ir : Bool .
\end{verbatim}
where the sort \verb|TaskID| is a supersort of \verb|MaybeNat| and
\verb|Oid|, referring to some task.
\begin{verbatim}
 subsorts MaybeNat Oid < TaskID .
\end{verbatim}
Some operations, such as \verb|getPc| and \verb|setMask|, are defined
on the class \verb|Regs|.

Then the hardware is described by the sort \verb|Hardware|:
\begin{verbatim}
 op [_;_] : Object Stack 
              ~> Hardware [ctor] .
 mb ([ < O:Oid : Regs |> ; S:Stack ]) 
      : Hardware .
\end{verbatim}

@@

\subsubsection{System}
The entire system in our model is a composition of the parts
introduced above, of a sort \verb|System|\footnote{Some variable
  declarations are not shown for simplicity.}:
\begin{verbatim}
 op _____ : TaskList Timer 
            SysTasks Hardware Object 
              ~> System [ctor] .
 mb (L T STS HW < O:Oid : IntSrc |>) 
      : System .
\end{verbatim}

\subsection{Interrupt Requests}
\label{ss:ir}
Interrupt requests are performed by the source exactly every cycle
$T$, when the attribute \verb|val| decreases to \verb|zero|. The
requesting is an instantaneous action, thus is modeled by the
following instantaneous conditional rule applied on \verb|System|:
\begin{verbatim}
 crl [interrupt-request] :
   (L T STS HW ISRC) 
   => (L T STS (HW).intReq reset(ISRC))
   if (ISRC).timeout .
\end{verbatim}
where the function \verb|_.timeout| examines whether the attribute
\verb|val| equals \verb|zero|, and \verb|_.intReq| sets the \verb|ir|
bit indicating there exists an interrupt request to be handled:
\begin{verbatim}
 op _.intReq : Hardware -> Hardware .
 eq [ REGS ; S ].intReq 
    = [ (REGS).setIr ; S ] .
\end{verbatim}
Then the request will wait to be handled, which is explained in
Section~\ref{ss:inthandling}.

@@

\subsection{Interrupt Handling and Task Scheduling}
\label{ss:inthandling}
When an interrupt request occurs, it may not be detected immediately
by the system. It requires the \verb|mask| bit to be cleared. Once the
request is detected, it is handled in two steps: the interrupt
handling mechanism of the hardware (such as clearing \verb|ir|,
pushing context into stack and so on), and to invoke the $schedule()$
function. This behavior is modeled by the following instantaneous
rewrite rule:
\begin{verbatim}
 crl [interrupt-handle] :
   SYSTEM 
   => ((SYSTEM).interrupt).startScheduling
   if (SYSTEM).existInt .
\end{verbatim}
where \verb|_.existInt| checks whether \verb|mask| is cleared
\emph{and} \verb|ir| is set. The function \verb|_.interrupt| models
the interrupt handling mechanism performed by the hardware and does
four things: (i) clearing the \verb|ir| bit, which means the request
has been handled; (ii) pushing the current \verb|pc| value into the
stack, storing the interrupted context; (iii) assigning
\verb|scheduling| of sort \verb|Oid| to \verb|pc|, which indicates
that the system is scheduling; and (iv) setting the \verb|mask| bit,
to mask coming interrupt requests.

Unlike periodic tasks, even though the scheduling stage is modeled by
a \verb|Counter|, its functionality is too important to be abandoned.
We divide the behaviors of scheduling into three parts.  The first
part contains its timed behaviors. This part is modeled by regarding
scheduling as a system task of sort \verb|SysTask|. Modeling timed
behaviors of tasks is explained in Section~\ref{ss:timedbehavior}.
The other two parts together define its functionality. The second part
corresponds to Lines~\ref{l:updatestatus}-\ref{l:inc} in
Figure~\ref{a:schedule}. It updates the status of $taskList$ and
increases $timer$ by $1$. This part is modeled by function
\verb|_.startScheduling| which applies instantaneously at the
beginning of scheduling, as shown in rule \verb|interrupt-handle|:
\begin{verbatim}
 op _.startScheduling : System -> System .
 eq (L T STS HW ISRC).startScheduling 
    = ((updateStatus L with T) 
       inc(T) STS HW ISRC) .
\end{verbatim}
The third part corresponds to
Lines~\ref{l:startrun1st}-\ref{l:endrun1st}, finding the first
should-be-run periodic task and setting it to execute. It is modeled
by function \verb|_.finishScheduling|, which applies instantaneously
at the end of scheduling:
\begin{verbatim}
 op _.finishScheduling : System -> System .
 eq (L T STS HW ISRC).finishScheduling
    = (L T (finish scheduling in STS) 
       HW ISRC).run1stTask .
\end{verbatim}
where \verb|finish_in_| resets the counter of task \verb|scheduling|,
and \verb|_.run1stTask| models
Lines~\ref{l:startrun1st}-\ref{l:endrun1st} in
Figure~\ref{a:schedule}, searching the task with highest priority
that has status \verb|INTERRUPT| or \verb|READY| then performing an
\emph{interrupt return} or executing it, respectively.

@@

Similar to scheduling, the switching stage is divided into timed
behaviors of \verb|switching| and its functionality. \verb|switching|
starts when the execution time of the running periodic task reaches
its computation requirement, and finishes when its own execution time
does so. Two similar instantaneous rules are defined to model the
functionality of switching:
\begin{verbatim}
 crl [task-finish] :
   (L T STS HW ISRC) 
   => (SYSTEM).startSwitching
   if SYSTEM := (L T STS HW ISRC)
      /\ some N := (SYSTEM).running
      /\ some N mayFinish?in L .
 crl [switching-finish] :
   (L T STS HW ISRC) 
   => (SYSTEM).finishSwitching
   if SYSTEM := (L T STS HW ISRC)
      /\ (SYSTEM).running == switching
      /\ switching mayFinish?in STS .
\end{verbatim}

\subsection{Timed Behaviors of the System}
\label{ss:timedbehavior}
Timed behaviors of the system consist of two parts: the execution of
tasks and the execution of the interrupt source. Both are modeled
together by the following single ``standard'' tick
rule~\cite{DBLP:journals/entcs/OlveczkyM07a}:
\begin{verbatim}
 crl [tick]:
   {SYSTEM} 
   => {delta(SYSTEM, R)} in time R 
   if R le mte(SYSTEM) [nonexec] .
\end{verbatim}
where \verb|delta| defines the effects of time elapse on the system,
and \verb|mte| denotes the \emph{m}aximum amount of \emph{t}ime
allowed to \emph{e}lapse from the current state until an instantaneous
transition \emph{must} be performed. In fact, the core to model timed
behaviors is to define functions \verb|delta| and \verb|mte|. Notice
that the variable \verb|R| is \emph{continuous} with respect to the
specific time domain\footnote{Real-Time Maude contains built-in
  modules to define the time domain to be natural numbers and rational
  numbers, specifying \emph{discrete} time domains and \emph{dense}
  time domains, respectively.}  that we choose to instantiate our
model on, which is different from timed automata that discretize dense
time by defining ``clock region''.

@@

\verb|mte|, the maximum amount of time allowed to elapse, depends on
when the next instantaneous action must perform. Therefore, it is
decided by three arguments: the remaining time to complete the running
task, the remaining time to request the next interrupt, and whether or
not there exists an interrupt request detected for the moment:
\begin{verbatim}
 ceq mte(L T STS HW ISRC)
     = minimum( mteTask(ID, L),
                minimum( mteIS(ISRC), 
                         mteIr(HW)))
     if ID := (HW).getPc 
        /\ ID :: MaybeNat .
\end{verbatim}
where \verb|mteIr| returns \verb|zero| if there exists an interrupt
request detected in the system, or \verb|INF| which represents
\emph{infinity} otherwise. Again we do not show the case where
\verb|ID| is of sort \verb|Oid|, which is very similar. We should
point out that \verb|mteTask| computes the remaining time to reach the
maximum of the computation requirement of the task, since that is the
time at which a \verb|task-finish| transition \emph{must} happen.

@@

\subsection{Properties}
We consider two properties in this paper: schedulability and
correctness. By schedulability, we examine whether a given task set is
schedulable by the implementation. By correctness, we verify whether
the implementation schedules periodic tasks exactly with respect to
their priorities and periods.

To verify the schedulability of a given set of periodic tasks, we
define an atomic proposition \verb|taskTimeout| to hold if there
exists an \verb|error| appearing in $taskList$ of the current state:
\begin{verbatim}
 op taskTimeout : -> Prop [ctor] .
 eq {L T STS HW ISRC} |= taskTimeout 
    = containError(L) .
\end{verbatim}
where \verb|containError| returns \verb|true| if there is an object
\verb|error| existing in \verb|L|. Then our desired
property--schedulability--can be formalized as the temporal logic
formula: \verb|[](~taskTimeout)|. As the property is not
\emph{clocked}, given an initial state \verb|init|, the following
untimed model checking command returns \verb|true| if the
schedulability property holds with no time limit; otherwise a trace
showing a counterexample is provided:
\begin{verbatim}
  (mc init |=u [](~taskTimeout) .)
\end{verbatim}

Another objective is to verify the correctness of the implementation.
The atomic proposition \verb|correct| is hence defined to hold if the
running periodic task is the one requested to be executed with the
highest priority:
\begin{verbatim}
 op correct : -> Prop [ctor] .
 ceq {L T STS HW ISRC} |= correct
     = if ID :: MaybeNat 
       then shouldRun(ID, L)
       else true fi
     if ID := (HW).getPc .
\end{verbatim}
where \verb|shouldRun(ID, L)| returns \verb|true| if the task
identified by \verb|ID|, probably \verb|none|, is the one possessing
the highest priority among those whose status is \verb|READY|,
\verb|RUNNING| or \verb|INTERRUPT|. Note that in this paper, we do not
care the behaviors after some task misses its deadline. Therefore, the
correctness property is formalized by the temporal logic formula:
\verb|([]correct) \/ (correct U taskTimeout)|, and can be verified by
the following untimed model checking command provided an initial state
\verb|init|:
\begin{verbatim}
 (mc init |=u (correct U taskTimeout) 
              \/ ([]correct) .)
\end{verbatim}

\subsection{Scenarios}
\label{ss:results}
We use the following setting for our verification, which is from the 
statistics provided by our industrial partner:
\begin{itemize}
\item The interrupt cycle $T$ is $5ms$.
\item The scheduling time ranges from $15{\mu}s$ to $38{\mu}s$, and the
switching time ranges from $10{\mu}s$ to $20{\mu}s$.
\hide{\item The scheduling time ranges from $5{\mu}s$ to $9{\mu}s$, and the
switching time ranges from $2{\mu}s$ to $4{\mu}s$.}
\item The initial state is with empty stack, empty \verb|pc|, cleared 
\verb|mask| and cleared \verb|ir|.
\end{itemize}
